---
title: "Report"
author:
  - name: "Noor Fatima Moolji"
  - name: "Miriam Wu"
  - name: "Pei Sin Ang" 
date: today
format:
    pdf:
      toc: true
execute:
    warning: false
    message: false
---
## Introduction 
Residential safety is a key concern for communities, policymakers, and law enforcement agencies across Victoria. 

Property crime which includes burglary, theft and more represent a direct threat to the security and wellbeing of residents. 

Understanding the spatial distribution of property crime across Local Government Areas (LGAs) is essential for informing targeted crime prevention strategies. 

Victoria is divided into four police regions, each encompassing a diverse range of urban, suburban, and rural LGAs with varying socioeconomic profiles.

This report analyses residential property crime trends across Victorian LGAs from 2016 to 2025, using data sourced from Crime Statistics Agency Victoria. The analysis focuses on offence types directly relevant to residential safety, excluding commercial and white-collar crimes that do not reflect neighbourhood-level risk. 

The report identifies the safest and least safe LGAs in Victoria as of 2025 and examines how crime rates vary across police regions. Findings are intended to support evidence-based decision-making by local governments and Victoria Police. 

## Methodology 

```{r}
# Read crime data in Vic
library("readxl")

vic_crime_raw <- read_excel(
  "data/Data_Tables_LGA_Criminal_Incidents_Year_Ending_December_2025.xlsx",
  sheet = "Table 02"
)

police_region_raw <- read_excel("data/Data_Tables_LGA_Criminal_Incidents_Year_Ending_December_2025.xlsx", 
sheet = "Table 01") 
```

```{r}
# Data cleaning
library(tidyverse)
# Clean table 01 to get police region ready for joining 
police_region_clean <- police_region_raw |>
  mutate(LGA = str_trim(`Local Government Area`)) |> # Remove white space from raw data
  filter(!LGA %in% c("Total", "Unincorporated Vic",
                     "Justice Institutions and Immigration Facilities")) |> # Remove rows with invalid LGA 
  distinct(LGA, `Police Region`) |> 
  rename(police_region = `Police Region`)

# Clean table 02 
vic_crime_clean <- vic_crime_raw |> 
    mutate(Year = as.factor(Year),
    incidents_recorded = as.integer(`Incidents Recorded`), 
    LGA = str_trim(`Local Government Area`)) |>  # Remove white space from raw data
    select(-`Incidents Recorded`, -`Year ending`, -`Local Government Area`, -`Police Service Area`) |> 
    # Split the code and descriptions into two separate column
    mutate(
        division_code = as.factor(str_split_fixed(`Offence Division`, " ", 2)[, 1]),
        division_desc = str_split_fixed(`Offence Division`, " ", 2)[, 2],
        subdivision_code = as.factor(str_split_fixed(`Offence Subdivision`, " ", 2)[, 1]),
        subdivision_desc = str_split_fixed(`Offence Subdivision`, " ", 2)[, 2],
        subgroup_code = as.factor(str_split_fixed(`Offence Subgroup`, " ", 2)[, 1]),
        subgroup_desc = str_split_fixed(`Offence Subgroup`, " ", 2)[, 2]
    ) |> 
    select(-`Offence Division`, -`Offence Subdivision`, -`Offence Subgroup`) |> 
    rename(lga_rate_per_100_pop = `LGA Rate per 100,000 population`,
    psa_rate_per_100k_pop = `PSA Rate per 100,000 population`)
```

```{r}
# Join with police region, and remove all non-property related crime
property_crime_vic <- vic_crime_clean |>
  left_join(police_region_clean, by = "LGA") |>
  filter(division_code == "B",
        !subdivision_desc == "Deception", 
        !str_detect(subgroup_desc, "Non-residential"),
        !str_detect(subgroup_desc, "retail"),
        !subgroup_code %in% c("B12", "B61", "B46", "B45", "B319")
        )

write_csv(property_crime_vic, "data/property_crime_vic.csv")
```

```{r}
#| include: false
# Sanity check
property_crime_vic|> 
    distinct(subgroup_code, subgroup_desc)
```

```{r}
#| label: tbl-region-summary
#| tbl-cap: "Number of LGAs by police region, Victoria"

property_crime_vic |>
  group_by(police_region) |>
  summarise(n_lgas = n_distinct(LGA)) |>
  knitr::kable(
    col.names = c("Police Region", "Number of LGAs")
  )
```

Use @tbl-tbl-region-summary to expand on what each police region represents geographically. 
https://www.police.vic.gov.au/victoria-police-corporate-plan-2023-2024/about-victoria-police

```{r}
#| label: fig-incidents-offence
#| fig-cap: "Property Crime Incidents by Offence Type (2016-2025)"

property_crime_vic |>
  group_by(Year, subdivision_desc) |>
  summarise(total = sum(incidents_recorded),
            .groups = "drop") |>
  ggplot(aes(x = Year, y = total, fill = subdivision_desc)) +
  geom_col(position = "stack") +
  labs(title = "Property Crime Incidents by Offence Type (2016–2025)",
       y = "Total Incidents", fill = "Offence Type") +
  theme_minimal()
```

@fig-incidents-offence
for methodology: can reference this and write about the 4 offence subdivision of property crime and the general trend over time
Can also be used for results if needed 

## Results

```{r}
#| label: fig-lga-ranking
#| fig-cap: "Top 10 Safest and Least Safe LGAs by Total Property Crime Rate per 100,000 Population (2025)"

lga_total_rate_2025 <- property_crime_vic |>
  filter(Year == 2025) |>
  group_by(LGA, police_region) |>
  summarise(total_rate = sum(lga_rate_per_100_pop), .groups = "drop")

top10_unsafe <- lga_total_rate_2025 |> slice_max(total_rate, n = 10) |> mutate(rank = "Least Safe")
top10_safe   <- lga_total_rate_2025 |> slice_min(total_rate, n = 10) |> mutate(rank = "Safest")

bind_rows(top10_unsafe, top10_safe) |>
  mutate(LGA = reorder(LGA, total_rate)) |>
  ggplot(aes(x = LGA, y = total_rate, fill = rank)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Least Safe" = "#d73027", "Safest" = "#4575b4")) +
  labs(
    x = "LGA", 
    y = "Total Property Crime Rate per 100,000 Population",
    fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

The results reveal considerable variation in residential property crime rates across Victorian LGAs in 2025. As shown in @fig-lga-ranking, the ten least safe LGAs recorded substantially higher total property crime rates per 100,000 population compared to the ten safest LGAs. Melbourne had the highest rate at approximately 8,500 incidents per 100,000 population, which is consistent with its status as a high-density urban centre. 

Notably, the ten safest LGAs were predominantly rural and regional areas, with Queenscliffe recording the lowest rate. This urban-rural divide suggests that population density and urbanisation may be key drivers of residential property crime risk in Victoria, and highlights the importance of targeted policy responses at the local government level.

Ideas: 
can rank safest and least safe LGA - useful for conclusion
also accidentally did this chart below, but not sure if it will be useful in results feel free to make edits. 
```{r}
#| label: fig-region-offence
#| fig-cap: "Average LGA Property Crime Rate by Police Region and Offence Type (2025)"
property_crime_vic |>
  filter(Year == 2025) |>
  group_by(LGA, police_region, subdivision_desc) |>
  # sum within each LGAL: add up subgroup lga rate 
  summarise(lga_subdivision_rate = sum(lga_rate_per_100_pop), .groups = "drop") |>
  group_by(police_region, subdivision_desc) |>
  # average across lgas within each police region
  summarise(mean_rate = mean(lga_subdivision_rate), .groups = "drop") |>
  ggplot(aes(x = police_region, y = mean_rate, fill = subdivision_desc)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average LGA Property Crime Rate by Police Region and Offence Type (2025)",
    x = "Police Region", y = "Mean Rate per 100,000 Population",
    fill = "Offence Type"
  ) +
  theme_minimal() 
```

## References 
(to be updated in reference.bib)
crime data download - https://www.crimestatistics.vic.gov.au/crime-statistics/latest-crime-data-by-area