---
title: "Report"
author:
  - name: "Noor Fatima Moolji"
  - name: "Miriam Wu"
  - name: "Pei Sin Ang" 
date: today
format:
    pdf:
      toc: true
execute:
    warning: false
    message: false
---
## Introduction 
topic: Assessing Residential Safety in Victoria: A Property Crime Analysis by Police Region and LGA (2016–2025)

## Methodology 

```{r}
# Read crime data in Vic
library("readxl")

vic_crime_raw <- read_excel(
  "data/Data_Tables_LGA_Criminal_Incidents_Year_Ending_December_2025.xlsx",
  sheet = "Table 02"
)

police_region_raw <- read_excel("data/Data_Tables_LGA_Criminal_Incidents_Year_Ending_December_2025.xlsx", 
sheet = "Table 01") 
```

```{r}
# Data cleaning
library(tidyverse)
# Clean table 01 to get police region ready for joining 
police_region_clean <- police_region_raw |>
  mutate(LGA = str_trim(`Local Government Area`)) |> # Remove white space from raw data
  filter(!LGA %in% c("Total", "Unincorporated Vic",
                     "Justice Institutions and Immigration Facilities")) |> # Remove rows with invalid LGA 
  distinct(LGA, `Police Region`) |> 
  rename(police_region = `Police Region`)

# Clean table 02 
vic_crime_clean <- vic_crime_raw |> 
    mutate(Year = as.factor(Year),
    incidents_recorded = as.integer(`Incidents Recorded`), 
    LGA = str_trim(`Local Government Area`)) |>  # Remove white space from raw data
    select(-`Incidents Recorded`, -`Year ending`, -`Local Government Area`, -`Police Service Area`) |> 
    # Split the code and descriptions into two separate column
    mutate(
        division_code = as.factor(str_split_fixed(`Offence Division`, " ", 2)[, 1]),
        division_desc = str_split_fixed(`Offence Division`, " ", 2)[, 2],
        subdivision_code = as.factor(str_split_fixed(`Offence Subdivision`, " ", 2)[, 1]),
        subdivision_desc = str_split_fixed(`Offence Subdivision`, " ", 2)[, 2],
        subgroup_code = as.factor(str_split_fixed(`Offence Subgroup`, " ", 2)[, 1]),
        subgroup_desc = str_split_fixed(`Offence Subgroup`, " ", 2)[, 2]
    ) |> 
    select(-`Offence Division`, -`Offence Subdivision`, -`Offence Subgroup`) |> 
    rename(lga_rate_per_100_pop = `LGA Rate per 100,000 population`,
    psa_rate_per_100k_pop = `PSA Rate per 100,000 population`)
```

```{r}
# Join with police region, and remove all non-property related crime
property_crime_vic <- vic_crime_clean |>
  left_join(police_region_clean, by = "LGA") |>
  filter(division_code == "B",
        !subdivision_desc == "Deception", 
        !str_detect(subgroup_desc, "Non-residential"),
        !str_detect(subgroup_desc, "retail"),
        !subgroup_code %in% c("B12", "B61", "B46", "B45", "B319")
        )

write_csv(property_crime_vic, "data/property_crime_vic.csv")
```

```{r}
#| include: false
# Sanity check
property_crime_vic|> 
    distinct(subgroup_code, subgroup_desc)
```

```{r}
#| label: tbl-region-summary
#| tbl-cap: "Number of LGAs by police region, Victoria"

property_crime_vic |>
  group_by(police_region) |>
  summarise(n_lgas = n_distinct(LGA)) |>
  knitr::kable(
    col.names = c("Police Region", "Number of LGAs")
  )
```

Use @tbl-tbl-region-summary to expand on what each police region represents geographically. 
https://www.police.vic.gov.au/victoria-police-corporate-plan-2023-2024/about-victoria-police

```{r}
#| label: fig-incidents-offence
#| fig-cap: "Property Crime Incidents by Offence Type (2016-2025)"

property_crime_vic |>
  group_by(Year, subdivision_desc) |>
  summarise(total = sum(incidents_recorded),
            .groups = "drop") |>
  ggplot(aes(x = Year, y = total, fill = subdivision_desc)) +
  geom_col(position = "stack") +
  labs(title = "Property Crime Incidents by Offence Type (2016–2025)",
       y = "Total Incidents", fill = "Offence Type") +
  theme_minimal()
```

@fig-incidents-offence
for methodology: can reference this and write about the 4 offence subdivision of property crime and the general trend over time
Can also be used for results if needed 

## Results

Ideas: 
can rank safest and least safe LGA - useful for conclusion
also accidentally did this chart below, but not sure if it will be useful in results feel free to make edits. 
```{r}
#| label: fig-region-offence
#| fig-cap: "Average LGA Property Crime Rate by Police Region and Offence Type (2025)"
property_crime_vic |>
  filter(Year == 2025) |>
  group_by(LGA, police_region, subdivision_desc) |>
  # sum within each LGAL: add up subgroup lga rate 
  summarise(lga_subdivision_rate = sum(lga_rate_per_100_pop), .groups = "drop") |>
  group_by(police_region, subdivision_desc) |>
  # average across lgas within each police region
  summarise(mean_rate = mean(lga_subdivision_rate), .groups = "drop") |>
  ggplot(aes(x = police_region, y = mean_rate, fill = subdivision_desc)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average LGA Property Crime Rate by Police Region and Offence Type (2025)",
    x = "Police Region", y = "Mean Rate per 100,000 Population",
    fill = "Offence Type"
  ) +
  theme_minimal() 
```

## References 
(to be updated in reference.bib)
crime data download - https://www.crimestatistics.vic.gov.au/crime-statistics/latest-crime-data-by-area